---
title: "DATA101 Week10"
author: "Abdirisak Mohamed"
date: "Spring 2018"
output:
  html_document:
    toc: yes
    toc_depth: '2'
  pdf_document:
    keep_tex: yes
    toc: yes
    toc_depth: 2
geometry: right=2.5in
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
url:
  lib: ../../librariesNew
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---

## Reference: Statistical Inference for Data Science, Brian Caffo



## P-values

* Most common measure of statistical significance
* Their ubiquity, along with concern over their interpretation and use
  makes them controversial among statisticians
  * [http://warnercnr.colostate.edu/~anderson/thompson1.html](http://warnercnr.colostate.edu/~anderson/thompson1.html)
  * Also see *Statistical Evidence: A Likelihood Paradigm* by Richard Royall 
  * *Toward Evidence-Based Medical Statistics. 1: The P Value Fallacy* by Steve Goodman
  * The hilariously titled: *The Earth is Round (p < .05)* by Cohen.
* Some positive comments
  * [simply statistics](http://simplystatistics.org/2012/01/06/p-values-and-hypothesis-testing-get-a-bad-rap-but-we/)
  * [normal deviate](http://normaldeviate.wordpress.com/2013/03/14/double-misunderstandings-about-p-values/)
  * [Error statistics](http://errorstatistics.com/2013/06/14/p-values-cant-be-trusted-except-when-used-to-argue-that-p-values-cant-be-trusted/)




## What is a P-value? 

__Idea__: Suppose nothing is going on - how unusual is it to see the estimate we got?

__Approach__: 

1. Define the hypothetical distribution of a data summary (statistic) when "nothing is going on" (_null hypothesis_)
2. Calculate the summary/statistic with the data we have (_test statistic_)
3. Compare what we calculated to our hypothetical distribution and see if the value is "extreme" (_p-value_)

---

## P-values
* The P-value is the probability under the null hypothesis of obtaining evidence as extreme or more extreme than that obtained
* If the P-value is small, then either $H_0$ is true and we have observed a rare event or $H_0$ is false
*  Suppos that you get a $T$ statistic of $2.5$ for 15 df testing $H_0:\mu = \mu_0$
versus $H_a : \mu > \mu_0$. 
  * What's the probability of getting a $T$ statistic as large as $2.5$?
```{r}
pt(2.5, 15, lower.tail = FALSE) 
```
* Therefore, the probability of seeing evidence as extreme or more extreme than that actually obtained under $H_0$ is `r pt(2.5, 15, lower.tail = FALSE)`

---

## The attained significance level
* Our test statistic was $2$ for $H_0 : \mu_0  = 30$ versus $H_a:\mu > 30$.
* Notice that we rejected the one sided test when $\alpha = 0.05$, would we reject if $\alpha = 0.01$, how about $0.001$?
* The smallest value for alpha that you still reject the null hypothesis is called the *attained significance level*
* This is equivalent, but philosophically a little different from, the *P-value*

---

## Notes
* By reporting a P-value the reader can perform the hypothesis
  test at whatever $\alpha$ level he or she choses
* If the P-value is less than $\alpha$ you reject the null hypothesis 
* For two sided hypothesis test, double the smaller of the two one
  sided hypothesis test Pvalues

---

## Revisiting an earlier example
- Suppose a friend has $8$ children, $7$ of which are girls and none are twins
- If each gender has an independent $50$% probability for each birth, what's the probability of getting $7$ or more girls out of $8$ births?
```{r}
choose(8, 7) * .5 ^ 8 + choose(8, 8) * .5 ^ 8 
pbinom(6, size = 8, prob = .5, lower.tail = FALSE)
```

---

## Poisson example
- Suppose that a hospital has an infection rate of 10 infections per 100 person/days at risk (rate of 0.1) during the last monitoring period.
- Assume that an infection rate of 0.05 is an important benchmark. 
- Given the model, could the observed rate being larger than 0.05 be attributed to chance?
- Under $H_0: \lambda = 0.05$ so that $\lambda_0 100 = 5$
- Consider $H_a: \lambda > 0.05$.

```{r}
ppois(9, 5, lower.tail = FALSE)
```


## Power
- Power is the probability of rejecting the null hypothesis when it is false
- Ergo, power (as its name would suggest) is a good thing; you want more power
- A type II error (a bad thing, as its name would suggest) is failing to reject the null hypothesis when it's false; the probability of a type II error is usually called $\beta$
- Note Power  $= 1 - \beta$

---

## Notes
- Consider our previous example involving RDI
- $H_0: \mu = 30$ versus $H_a: \mu > 30$
- Then power is 
$$P\left(\frac{\bar X - 30}{s /\sqrt{n}} > t_{1-\alpha,n-1} ~;~ \mu = \mu_a \right)$$
- Note that this is a function that depends on the specific value of $\mu_a$!
- Notice as $\mu_a$ approaches $30$ the power approaches $\alpha$


---

## Calculating power for Gaussian data
- We reject if $\frac{\bar X - 30}{\sigma /\sqrt{n}} > z_{1-\alpha}$    
    - Equivalently if $\bar X > 30 + Z_{1-\alpha} \frac{\sigma}{\sqrt{n}}$
- Under $H_0 : \bar X \sim N(\mu_0, \sigma^2 / n)$
- Under $H_a : \bar X \sim N(\mu_a, \sigma^2 / n)$
- So we want 
```{r, echo=TRUE,eval=FALSE}
alpha = 0.05
z = qnorm(1 - alpha)
pnorm(mu0 + z * sigma / sqrt(n), mean = mua, sd = sigma / sqrt(n), 
      lower.tail = FALSE)
```

---

## Example continued
- $\mu_a = 32$, $\mu_0 = 30$, $n =16$, $\sigma = 4$
```{r, echo=TRUE,eval=TRUE}
mu0 = 30; mua = 32; sigma = 4; n = 16
alpha = 0.05
z = qnorm(1 - alpha)
pnorm(mu0 + z * sigma / sqrt(n), mean = mu0, sd = sigma / sqrt(n), 
      lower.tail = FALSE)
pnorm(mu0 + z * sigma / sqrt(n), mean = mua, sd = sigma / sqrt(n), 
      lower.tail = FALSE)
```

---

##  Plotting the power curve

```{r, fig.align='center', fig.height=6, fig.width=12, echo=FALSE}
library(ggplot2)
nseq = c(8, 16, 32, 64, 128)
mua = seq(30, 35, by = 0.1)
z = qnorm(.95)
power = sapply(nseq, function(n)
pnorm(mu0 + z * sigma / sqrt(n), mean = mua, sd = sigma / sqrt(n), 
          lower.tail = FALSE)
    )
colnames(power) <- paste("n", nseq, sep = "")
d <- data.frame(mua, power)
library(reshape2)
d2 <- melt(d, id.vars = "mua")
names(d2) <- c("mua", "n", "power")    
g <- ggplot(d2, 
            aes(x = mua, y = power, col = n)) + geom_line(size = 2)
g            
```


---

## Graphical Depiction of Power
```{r, echo = TRUE, eval=FALSE}
library(manipulate)
mu0 = 30
myplot <- function(sigma, mua, n, alpha){
    g = ggplot(data.frame(mu = c(27, 36)), aes(x = mu))
    g = g + stat_function(fun=dnorm, geom = "line", 
                          args = list(mean = mu0, sd = sigma / sqrt(n)), 
                          size = 2, col = "red")
    g = g + stat_function(fun=dnorm, geom = "line", 
                          args = list(mean = mua, sd = sigma / sqrt(n)), 
                          size = 2, col = "blue")
    xitc = mu0 + qnorm(1 - alpha) * sigma / sqrt(n)
    g = g + geom_vline(xintercept=xitc, size = 3)
    g
}
manipulate(
    myplot(sigma, mua, n, alpha),
    sigma = slider(1, 10, step = 1, initial = 4),
    mua = slider(30, 35, step = 1, initial = 32),
    n = slider(1, 50, step = 1, initial = 16),
    alpha = slider(0.01, 0.1, step = 0.01, initial = 0.05)
    )

```


---

## Question
- When testing $H_a : \mu > \mu_0$, notice if power is $1 - \beta$, then 
$$1 - \beta = P\left(\bar X > \mu_0 + z_{1-\alpha} \frac{\sigma}{\sqrt{n}} ; \mu = \mu_a \right)$$
- where $\bar X \sim N(\mu_a, \sigma^2 / n)$
- Unknowns: $\mu_a$, $\sigma$, $n$, $\beta$
- Knowns: $\mu_0$, $\alpha$
- Specify any 3 of the unknowns and you can solve for the remainder

---

## Notes
- The calculation for $H_a:\mu < \mu_0$ is similar
- For $H_a: \mu \neq \mu_0$ calculate the one sided power using
  $\alpha / 2$ (this is only approximately right, it excludes the probability of
  getting a large TS in the opposite direction of the truth)
- Power goes up as $\alpha$ gets larger
- Power of a one sided test is greater than the power of the
  associated two sided test
- Power goes up as $\mu_1$ gets further away from $\mu_0$
- Power goes up as $n$ goes up
- Power doesn't need $\mu_a$, $\sigma$ and $n$, instead only $\frac{\sqrt{n}(\mu_a - \mu_0)}{\sigma}$
  - The quantity $\frac{\mu_a - \mu_0}{\sigma}$ is called the effect size, the difference in the means in standard deviation units.
  - Being unit free, it has some hope of interpretability across settings

---

## T-test power
-  Consider calculating power for a Gossett's $T$ test for our example
-  The power is
  $$
  P\left(\frac{\bar X - \mu_0}{S /\sqrt{n}} > t_{1-\alpha, n-1} ~;~ \mu = \mu_a \right)
  $$
- Calcuting this requires the non-central t distribution.
- `power.t.test` does this very well
  - Omit one of the arguments and it solves for it

---

## Example
```{r}
power.t.test(n = 16, delta = 2 / 4, sd=1, type = "one.sample",  alt = "one.sided")$power
power.t.test(n = 16, delta = 2, sd=4, type = "one.sample",  alt = "one.sided")$power
power.t.test(n = 16, delta = 100, sd=200, type = "one.sample", alt = "one.sided")$power
```

---

## Example
```{r}
power.t.test(power = .8, delta = 2 / 4, sd=1, type = "one.sample",  alt = "one.sided")$n
power.t.test(power = .8, delta = 2, sd=4, type = "one.sample",  alt = "one.sided")$n
power.t.test(power = .8, delta = 100, sd=200, type = "one.sample", alt = "one.sided")$n
```

## Key ideas

* Hypothesis testing/significance analysis is commonly overused
* Correcting for multiple testing avoids false positives or discoveries
* Two key components
  * Error measure
  * Correction


---

## Three eras of statistics

__The age of Quetelet and his successors, in which huge census-level data sets were brought to bear on simple but important questions__: Are there more male than female births? Is the rate of insanity rising?

The classical period of Pearson, Fisher, Neyman, Hotelling, and their successors, intellectual giants who __developed a theory of optimal inference capable of wringing every drop of information out of a scientific experiment__. The questions dealt with still tended to be simple Is treatment A better than treatment B? 

__The era of scientific mass production__, in which new technologies typified by the microarray allow a single team of scientists to produce data sets of a size Quetelet would envy. But now the flood of data is accompanied by a deluge of questions, perhaps thousands of estimates or hypothesis tests that the statistician is charged with answering together; not at all what the classical masters had in mind. Which variables matter among the thousands measured? How do you relate unrelated information?

[http://www-stat.stanford.edu/~ckirby/brad/papers/2010LSIexcerpt.pdf](http://www-stat.stanford.edu/~ckirby/brad/papers/2010LSIexcerpt.pdf)



## Reasons for multiple testing






## Why correct for multiple tests?




##[http://xkcd.com/882/](http://xkcd.com/882/)



## Why correct for multiple tests?



##[http://xkcd.com/882/](http://xkcd.com/882/)


