---
title: "Project 2 - MC DATA 101"
author: "Juan J. Nunez"
date: "April 14, 2018"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




First I'm going to start by turning an SPSS file (i.e., .sav) into an RMarkdown file (i.e., .Rda). The data I will be using is from the World Bank. For more information on how to download World Bank data, please visit https://data.worldbank.org

Once the data is downloaded to a .SAV file, it is easy to use it using R by bringing up the 'foreign' package.



```{r}
setwd("C:\\Users\\Juan Nunez\\Desktop\\MC_DATA_101\\ASSIG_2_DATA101")

```



I can look at the data using 'ggplot'. I download and bring up the package. First I get the data.



```{r}
##install.packages("ggplot2")
library(ggplot2)
```



Now I look at the dimensions of the ASSIG2_DATA data frame.



```{r}
library(foreign)
NEWLES <- read.spss("C:\\Users\\Juan Nunez\\Desktop\\MC_DATA_101\\ASSIG_2_DATA101\\LES101.SAV", use.value.labels=TRUE, to.data.frame=TRUE)

ASSIG2_DATA <- NEWLES

save(ASSIG2_DATA, file = "ASSIG2_DATA.Rda")

dim(ASSIG2_DATA)
```



I see that there are 148 rows and 356 columns. Let's look at a basic scatterplot first.



```{r}
ggplot(ASSIG2_DATA, aes(V14, V168)) +
  geom_point()
```



What we can see here is that there is a trend where as V14 increases, V168 decreases. To understand this better, we use the codebook below to understand what our variables of interest mean.

V1 COUNTRY NUMBER ; 
V2 ABBREVIATED COUNTRY NAME ; 
V3 COUNTRY NAME ; 
V5 % ADULT FEMALE ILLITERACY 1990 ; 
V12 ENERGY CONSUMPTION/CAPITA 1991 ; 
V14 INFANT MORTALITY RATE 1991 ; 
V168 FEMALE SECODARY SCHOOL ENROLLMENT GROSS 1980 ; 
V133 CIVIL LIBERTIES 1991 ; 
V188 WORLD AS 5 REGIONS ;




```{r}
library(dplyr)
SHORT_ASSIG2_DATA <- select(ASSIG2_DATA, V1, V2, V3, V5, V12, V14, V168, V133)
```

There is another issue that we have to take care of before we move on. Are there any missing values in V14 or V168?


```{r}
is.na(ASSIG2_DATA$V168)
summary(ASSIG2_DATA$V14)
summary(ASSIG2_DATA$V168)
is.na(ASSIG2_DATA$V14)
```

Apparently there are some missing values. So we should take a look at some statistics. Let's look at the variance, standard deviation, and inner quartile range.


```{r}

class(SHORT_ASSIG2_DATA)

class(SHORT_ASSIG2_DATA$V14)

var(SHORT_ASSIG2_DATA$V14, na.rm = TRUE)
sd(SHORT_ASSIG2_DATA$V14, na.rm = TRUE)
IQR(SHORT_ASSIG2_DATA$V14, na.rm = TRUE)
var(ASSIG2_DATA$V168, na.rm = TRUE)
sd(ASSIG2_DATA$V168, na.rm = TRUE)
IQR(ASSIG2_DATA$V168, na.rm = TRUE)
```


What about the mode?


```{r}

vect2 <- na.omit(SHORT_ASSIG2_DATA$V168)

getmode <- function(v) {
   uniqv <- unique(v)
   uniqv[which.max(tabulate(match(v, uniqv)))]
}

v <- vect2


result <- getmode(v)
print(result)


vect3 <- na.omit(SHORT_ASSIG2_DATA$V14)

getmode2 <- function(z) {
   uniqz <- unique(z)
   uniqz[which.max(tabulate(match(z, uniqv)))]
}

z <- vect3

result2 <- getmode(z)
print(result2)
```



Now that we have seen the statistics, we can start to look at the variables graphically.



```{r}
 boxplot(vect2, col = "blue")
 boxplot(vect3, col = "purple")
 hist(vect2, col = "green")
 hist(vect3,col = "red")
 table(vect2) %>% barplot(col = "wheat")
 table(vect3) %>% barplot(col = "orange")
```


Let's try stacked barplots. We are going to use the variable civil liberties. We have to make sure to use listwise deletion for this to work.

```{r}




SMALL_DATA <- select(SHORT_ASSIG2_DATA, V14, V133) 

SMALL_DATA <- na.omit(SMALL_DATA)

SMALL_DATA <- mutate(SMALL_DATA, INF_DUM = as.numeric(SMALL_DATA$V14 >= mean(SMALL_DATA$V14))) 

head(SMALL_DATA)






counts <- table(SMALL_DATA$INF_DUM, SMALL_DATA$V133)
barplot(counts, main="Countries Infant Moratlity Rate by Civil Liberties",
  xlab="Civil Liberties", col=c("darkblue","red"),
  legend = rownames(counts))
```


These graphics are very helpful to identify trends in the data. Let's take a look at a scatterplot now.

```{r}

SCATDAT <- select(SHORT_ASSIG2_DATA, V14, V168)
SCATDAT <- na.omit(SCATDAT)


 with(SCATDAT, plot(V14, V168, xlab = "Infant Mortality Rate", ylab = "Female Education")) 
```


We can confirm the association that we observed at the beginning of this analysis. Sometimes statistical analyses are not linear. 

Now let's see if we can incorporate civil liberties.


```{r}

SCATDAT2 <- select(SHORT_ASSIG2_DATA, V14, V168, V133)
SCATDAT2 <- na.omit(SCATDAT2)
SCATDAT2 <- mutate(SCATDAT2, CLIBDUM = as.numeric(SCATDAT2$V133 > 3)) 

head(SCATDAT2)

 with(SCATDAT2, plot(V14, V168, col = as.factor(CLIBDUM) ) ) 
```


Finally, we can see how those with more than 3 in civil liberties are red. Now we'll use ggplot again.


```{r}
ggplot(SCATDAT) +
  aes(x = V14, y = V168) +
  geom_point()
```

aes stands for aesthetics. Now we are going to get a line instead of dots.


```{r}
ggplot(SCATDAT) +
  aes(x = V14, y = V168) +
  geom_line()
```


Let's add some layers to our previous graphs.


```{r}
ggplot(SCATDAT) +
  aes(x = V14, y = V168) +
  geom_point(colour = 'red', alpha = 0.2)
```


Now let's look at the three variables together.

```{r}
ggplot(SCATDAT2) +
  aes(x = V14, y = V168, colour = CLIBDUM) +
  geom_point()
```


We can use a variable with more categories.

```{r}
SCATDAT3 <- select(SHORT_ASSIG2_DATA, V14, V168, V133)
SCATDAT3 <- na.omit(SCATDAT3)
SCATDAT3 <- mutate(SCATDAT3, V133CAT = as.factor(SCATDAT2$V133))

head(SCATDAT3)

ggplot(SCATDAT3) +
  aes(x = V14, y = V168, colour = V133CAT) +
  geom_point()
```

We can even do what is called a facet. We are going to use region of the world.


```{r}
SCATDAT4 <- select(ASSIG2_DATA, V14, V168, V133, V188)
SCATDAT4 <- na.omit(SCATDAT4)
SCATDAT4 <- mutate(SCATDAT4, V133CAT = as.factor(SCATDAT4$V133))
SCATDAT4 <- mutate(SCATDAT4, V188CAT = as.factor(SCATDAT4$V188))

ggplot(SCATDAT4) +
  aes(x = V14, y = V168, colour = V133CAT) +
  geom_point() +
  facet_grid(~ V188CAT)
```

Now we have a plot for each world region. The codebook of V188 states that the regions are: 

1=Sub-Saharan Africa; 
2=South Asia, East Asia, and Pacific; 
3=Europe/Central Asia; 
4=Middle East and North Africa; 
5=Americas;

Now let's see what ggplot can do.

```{r}
ggplot(SCATDAT4) +
  aes(x = V188) +
  geom_bar()
```

Now we create a dummy variable for civil liberties where those with more than 3 in the scale are coded as 1.


```{r}
SCATDAT4 <- mutate(SCATDAT4, CLIBDUM = as.numeric(SCATDAT2$V133 > 3)) 
head(SCATDAT4)
SCATDAT4 %>%
ggplot() +
  aes(x = V133CAT, fill = V188CAT) +
  geom_bar()
```

This is a bit messy, so we try to ammeliorate it.

```{r}
SCATDAT4 <- mutate(SCATDAT4, CLIBDUM = as.numeric(SCATDAT2$V133 > 3)) 
SCATDAT4 <- mutate(SCATDAT4, clibcat = as.factor(SCATDAT4$CLIBDUM))
head(SCATDAT4)
SCATDAT4 %>%
ggplot() +
  aes(x = V188CAT, fill = clibcat) +
  geom_bar()
```

Much better. We can make it even better by changing the positions of the bars.

```{r}
SCATDAT4 <- mutate(SCATDAT4, CLIBDUM = as.numeric(SCATDAT2$V133 > 3)) 
SCATDAT4 <- mutate(SCATDAT4, clibcat = as.factor(SCATDAT4$CLIBDUM))
head(SCATDAT4)
SCATDAT4 %>%
ggplot() +
  aes(x = V188CAT, fill = clibcat) +
  geom_bar(position = 'dodge')
```

We can look at proportions of the categorical value also.

```{r}
SCATDAT4 <- mutate(SCATDAT4, CLIBDUM = as.numeric(SCATDAT2$V133 > 3)) 
SCATDAT4 <- mutate(SCATDAT4, clibcat = as.factor(SCATDAT4$CLIBDUM))
head(SCATDAT4)
SCATDAT4 %>%
ggplot() +
  aes(x = V188CAT, fill = clibcat) +
  geom_bar(position = 'fill')
```

We can see how civil liberties take over proportions in different regions. Now we are going to look at heat plot. This is a useful tool to look for interactions.

```{r}
HEATDAT <- 

SCATDAT4 %>%
  group_by(clibcat, V188CAT) %>%
  summarise(avg_IMR = mean(V14)) %>%
  ungroup()

ggplot(HEATDAT) +
  aes(y = clibcat, x = V188CAT, fill = avg_IMR) +
  geom_raster()

HEATDAT
```


We can see in table all of the combinations of civil liberty and region and their corresponding average infant mortality rate. Let's put the mean infant mortality in the graph.

```{r}
HEATDAT$label <- HEATDAT$avg_IMR %>% round(1) %>% as.character
HEATDAT
```

We can also add color schemes. We need to install and download RColorBrewer.



```{r}
##install.packages('RColorBrewer')
library(RColorBrewer)

ggplot(HEATDAT) + 
  aes(y = clibcat, x = V188CAT, fill = avg_IMR, label = label) +
  geom_raster() + 
  geom_text(colour = 'white',
            size = 6) + 
  scale_fill_distiller(name = 'Average Infant Mortality Rate',
                       type = 'div',
                       palette = 3) +
  xlab('Region') +
  ylab('clibcat') +
  theme_bw()
```



